Many claims of group polarization are plausibly false due to inappropriate statistics

### 2023-10-18 GroupPolarizationStatmod/paper/ms.tex
update paper abstract

\title{Many claims of group polarization are plausibly false due to inappropriate
    statistics}
% \shorttitle{Group polarization or simple consensus?}
\author[1,2]{{Matthew A.~Turner}}
\affil[1]{Department of Earth System Science, Stanford University}
\affil[2]{Division of Social Sciences, Stanford Doerr School of Sustainability, Stanford University}

\author[3,4,5]{{Paul E.~Smaldino}}
\affil[3]{Cognitive and Information Sciences, University of California, Merced}
\affil[4]{Center for Advanced Study in the Behavioral Sciences, Stanford University}
\affil[5]{Santa Fe Institute}

@@

\linenumbers
\modulolinenumbers[3]

\begin{abstract}
  \noindent
  Psychology is beset by three crises: the reproducibility crisis, the
  generalizability crisis, and the theory crisis. New verbal theories for various
  phenomena are constantly being developed orthogonal to other theories, all based on
  mostly unreproducible analyses; even if a positive result is in some sense ``true''
  it may not account for all the variance present in a set of observations---studies
  thus fail to generalize because their results do not account for this variance.
  There is another related problem: a phenomenon may be ``reproducible'' only
  because studies of the phenomenon systematically misuse statistics in such a way
  that induces false positives for each succeeding ``replication.'' Here we show that
  replicated detections of group polarization are often plausibly false due to
  treating ordinal measurements of opinions, e.g., on a Likert-type scale, as drawn
  directly from an underlying continuous latent opinion distribution, which masks
  extreme latent opinions. Our intensive review of sixty experimental conditions
  across ten foundational group polarization papers revealed that 95\% of the
  conditions generated plausibly false positive results. To show this, we developed a
  novel generative model that provides two levels of evidence that an historical
  published empirical finding of group polarization is plausibly false.  Our model
  can be used to evaluate additional studies in group polarization and beyond, even
  if the reported data and significant metrics are incompletely reported, and the
  original raw data has been lost to time.  This may be just the tip of the
  plausibly-false-positives-in-psychology iceberg, both for group polarization
  studies and for the legion psychological results that model ordinal observations of
  opinions, beliefs, emotions, or any other latent psychological variable that is
  theoretically continuous, but measured using an ordinal scale.  
\end{abstract}
\clearpage

>>>>

\title{Many claims of group polarization are plausibly false due to inappropriate
    statistics}
% \shorttitle{Group polarization or simple consensus?}
% \author[1,2]{{Matthew A.~Turner}}
\author[1]{{Matthew A.~Turner}}
% \affil[1]{Department of Earth System Science, Stanford University}
\affil[1]{Division of Social Sciences, Stanford Doerr School of Sustainability, Stanford University}

% \author[3,4,5]{{Paul E.~Smaldino}}
\author[2,3]{{Paul E.~Smaldino}}
\affil[2]{Cognitive and Information Sciences, University of California, Merced}
% \affil[4]{Center for Advanced Study in the Behavioral Sciences, Stanford University}
\affil[3]{Santa Fe Institute}

@@

\linenumbers
% \modulolinenumbers[3]

\begin{abstract}
  \noindent
  % Psychology is beset by three crises: the reproducibility crisis, the
  % generalizability crisis, and the theory crisis. New verbal theories for various
  % phenomena are constantly being developed orthogonal to other theories, all based on
  % mostly unreproducible analyses; even if a positive result is in some sense ``true''
  % it may not account for all the variance present in a set of observations---studies
  % thus fail to generalize because their results do not account for this variance.
  % There is another related problem: a phenomenon may be ``reproducible'' only
  % because studies of the phenomenon systematically misuse statistics in such a way
  % that induces false positives for each succeeding ``replication.'' 
  \emph{Group polarization} refers to the observation that like-minded novel groups tend to
  become more extreme in their opinions after discussion as group members found consensus.
  \emph{Consensus} here refers to decreasing opinion variance; we further define
  \emph{simple consensus} as a process where group opinion variance decreases but
  the mean opinion does not change.
  Here we show that many published experimental detections of group polarization are explained
  equally rigorously as \emph{simple consensus}. 
  The reports we analyze assumed in their statistical analyses 
  that both latent psychological opinions \emph{and} the ordinal 
  reported opinions (e.g., on a Likert-type scale) are continuous. We use
  a generative model to demonstrate that 95\% of sixty experimental conditions
  across ten foundational group polarization papers report plausibly false detections of
  group polarization. We show that simulated data from these plausibly false
  cases often fools a $t$-test into reporting significant group polarizaiton
  effects, but a Bayesian implementation of the more appropriate ordered-probit
  model reliably identifies these data as
  representing simple consensus, not group polarization.
  Our results demonstrate the need for a new wave of group polarization research
  that uses rigorous statistical methods appropriate for the study design. 
  Our approach can be used beyond group polarization
  research to analyze results from any experimental paradigm that may use statistical models that
  are poorly matched to the study design. Replications are worthless if the statistics are wrong.

  % To show this, we developed a
  % novel generative model that provides two levels of evidence that an historical
  % published empirical finding of group polarization is plausibly false.  Our model
  % can be used to evaluate additional studies in group polarization and beyond, even
  % if the reported data and significant metrics are incompletely reported, and the
  % original raw data has been lost to time.  This may be just the tip of the
  % plausibly-false-positives-in-psychology iceberg, both for group polarization
  % studies and for the legion psychological results that model ordinal observations of
  % opinions, beliefs, emotions, or any other latent psychological variable that is
  % theoretically continuous, but measured using an ordinal scale.  
\end{abstract}
\clearpage
